Conversations in time: interactive visualisation to explore structured temporal data

### 2020-09-13 tsibbletalk.Rmd
review interactivity

* An ensemble of graphics
* Accelerate the exploratory data visualization process

@@

The \CRANpkg{tsibble} package [@wang2020tsibble] introduces a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. That said, a `tsibble` extends the `data.frame` and \CRANpkg{tibble} class with temporally contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. Below displays the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package. The `Month` column holds year-months as `index`. The `State` together with `Industry` are the identifiers for these 152 series, highlighted as `key`. Note that the column `Series ID` could be an alternative option for setting up `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent temporal data analysis later.

@@

* [crossfilter.js](http://crossfilter.github.io/crossfilter/) & [dc.js](https://dc-js.github.io/dc.js/)
* {cranvas} and {cranvastime}
* {crosstalk} and html widgets
* {rJava}, {rbokeh} {loon}

@@

The \CRANpkg{tsibbletalk} package, inspired by the \CRANpkg{crosstalk} package, introduces a shared tsibble data structure on top of a `tsibble` to allow for frictionless communication between different plots for temporal data. The `as_shared_tsibble()` function provides an entry point in the integrated flow, turning a `tsibble` to a shared instance (i.e. `SharedTsibbleData` subclassing of `SharedData` from \CRANpkg{crosstalk}) that powers data transmission across multiple views. The \CRANpkg{tsibbletalk} package aims to streamline interactive graphical analysis with the focus of temporal and structured linking.

@@

In summary, shared tsibble data nicely bridges between the \CRANpkg{crosstalk} and **tidyverts** ecosystems. The `as_shared_tsibble()` provides a symbolic user interface for effortless construction of a hybrid of hierarchical and categorical linkings. And the `plotly_key_tree()` in turn decodes the specification to plot a tree for data overview and navigation, accompanied with more detailed plots.

@@

This shiny module, decoupled to `tsibbleDiceUI()` and `tsibbleDiceServer()`, presents a clean interface and forms a resusable piece in a shiny application. Like all shiny modules, the first argument in both functions requires a user-supplied session id that must be unique. The UI function `tsibbleDiceUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is certainly the server function `tsibbleDiceServer()`, encapsulating the core algorithm that transforms data and sends messages to update the plot accordingly. It expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, and thereby the time `index` can be retrieved for dissection. The `period` argument is to specify the minimum number of seasonal blocks as desired, for example data shifted by "1 day", "2 days", or "1 week", etc. The following code chunk generates Figure \ref{fig:wrap-ped}.

>>>>

* An ensemble of graphics
* Accelerate the exploratory data visualisation process

@@

The \CRANpkg{tsibble} package [@wang2020tsibble] introduces a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. That said, a `tsibble` extends the `data.frame` and \CRANpkg{tibble} classes with temporally contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. Below displays the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package. The `Month` column holds year-months as `index`. The `State` together with `Industry` are the identifiers for these 152 series, highlighted as `key`. Note that the column `Series ID` could be an alternative option for setting up `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent temporal data analysis later.

@@

The R community has a long-standing contribution to interactive visualisation systems in abundance. These systems can be roughly divided into two classes: ones with web technology and the others without.

R \CRANpkg{shiny} and \CRANpkg{htmlwidgets} lay a fundamental infrastructure in marrying R with HTML elements and Javascript for interactivity, although they do not emphasise on interactive graphics. The \CRANpkg{htmlwidgets} package opens a door to embedding Javascript libraries into R such that users are able to write only R code to generate web-based plots. Many Javascript charting libraries have been ported to R as HTML widgets, including \CRANpkg{plotly}, \CRANpkg{rbokeh}, and \CRANpkg{leaflet} for maps. To enable interactions between different widgets, it can be achieved with \CRANpkg{shiny} or \CRANpkg{crosstalk}. The \CRANpkg{crosstalk} extends \CRANpkg{htmlwidgets} with shared R6 instances to support linked brushing and filtering across widgets.

If not using web technology, two examples are provided here: \CRANpkg{loon} based on Tcl/Tk and \CRANpkg{cranvas} based on Qt. They offer a wide array of pre-defined interactions, such as selecting and zooming, to manipulate plots via mouse, keyboard strokes, and menus. The \CRANpkg{cranvastime} is an add-on of \CRANpkg{cranvas}, which develops a set of novel interactions for temporal data, for example wrapping and mirroring.

@@

The \CRANpkg{tsibbletalk} package, inspired by the \CRANpkg{crosstalk} package, introduces a shared tsibble instance on top of a `tsibble` to allow for frictionless communication between different plots for temporal data. The `as_shared_tsibble()` function provides an entry point in the integrated flow, turning a `tsibble` to a shared instance (i.e. `SharedTsibbleData` subclassing of `SharedData` from \CRANpkg{crosstalk}) that powers data transmission across multiple views. The \CRANpkg{tsibbletalk} package aims to streamline interactive graphical analysis with the focus of temporal and structured linking.

@@

In summary, shared tsibble data nicely bridges between the \CRANpkg{crosstalk} and **tidyverts** ecosystems for temporal data over their common term "key". The `as_shared_tsibble()` provides a symbolic user interface for effortless construction of a hybrid of hierarchical and categorical linkings. And the `plotly_key_tree()` in turn decodes the specification to plot a tree for data overview and navigation, accompanied with more detailed plots.

@@

This shiny module, decoupled to `tsibbleDiceUI()` and `tsibbleDiceServer()`, presents a clean interface and forms a resusable piece in a shiny application. A shiny module provides the vehicle in modularising shiny applications for both users and developers. Like all shiny modules, the first argument in both functions requires a user-supplied session id that must be unique. The UI function `tsibbleDiceUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is certainly the server function `tsibbleDiceServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, and thereby the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. As for date-times (represented by `POSIXt`), it ranges from from fine "day" to coarse "year". The following code snippet generates Figure \ref{fig:wrap-ped}.

### 2020-09-13 tsibbletalk.tex
review interactivity

\begin{itemize}
\tightlist
\item
  An ensemble of graphics
\item
  Accelerate the exploratory data visualization process
\end{itemize}

@@

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduces a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. That said, a \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} class with temporally
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns.
Below displays the monthly Australian retail trade turnover data
(\texttt{aus\_retail}), available in the \CRANpkg{tsibbledata} package.
The \texttt{Month} column holds year-months as \texttt{index}. The
\texttt{State} together with \texttt{Industry} are the identifiers for
these 152 series, highlighted as \texttt{key}. Note that the column
\texttt{Series\ ID} could be an alternative option for setting up
\texttt{key}, but \texttt{State} and \texttt{Industry} are more readable
and informative. The \texttt{index} and \texttt{key} are ``sticky''
columns to a \texttt{tsibble}, forming critical pieces for fluent
temporal data analysis later.

@@

\begin{itemize}
\tightlist
\item
  \{cranvas\} and \{cranvastime\}
\item
  \href{http://crossfilter.github.io/crossfilter/}{crossfilter.js} \&
  \href{https://dc-js.github.io/dc.js/}{dc.js}
\item
  \{crosstalk\} and html widgets
\item
  \{rJava\}, \{rbokeh\} \{loon\}
\end{itemize}

@@

The \CRANpkg{tsibbletalk} package, inspired by the \CRANpkg{crosstalk}
package, introduces a shared tsibble data structure on top of a
\texttt{tsibble} to allow for frictionless communication between
different plots for temporal data. The \texttt{as\_shared\_tsibble()}
function provides an entry point in the integrated flow, turning a
\texttt{tsibble} to a shared instance (i.e.~\texttt{SharedTsibbleData}
subclassing of \texttt{SharedData} from \CRANpkg{crosstalk}) that powers
data transmission across multiple views. The \CRANpkg{tsibbletalk}
package aims to streamline interactive graphical analysis with the focus
of temporal and structured linking.

@@

In summary, shared tsibble data nicely bridges between the
\CRANpkg{crosstalk} and \textbf{tidyverts} ecosystems. The
\texttt{as\_shared\_tsibble()} provides a symbolic user interface for
effortless construction of a hybrid of hierarchical and categorical
linkings. And the \texttt{plotly\_key\_tree()} in turn decodes the
specification to plot a tree for data overview and navigation,
accompanied with more detailed plots.

@@

This shiny module, decoupled to \texttt{tsibbleDiceUI()} and
\texttt{tsibbleDiceServer()}, presents a clean interface and forms a
resusable piece in a shiny application. Like all shiny modules, the
first argument in both functions requires a user-supplied session id
that must be unique. The UI function \texttt{tsibbleDiceUI()} simply
shows a slider that animates or controls the number of periods to be
diced. The workhorse is certainly the server function
\texttt{tsibbleDiceServer()}, encapsulating the core algorithm that
transforms data and sends messages to update the plot accordingly. It
expects a \texttt{ggplot} or \texttt{plotly} object, where one can plot
data using either lines or other graphical elements (such as boxplots).
As the function name suggests, a (shared) tsibble is needed to start the
engine, and thereby the time \texttt{index} can be retrieved for
dissection. The \texttt{period} argument is to specify the minimum
number of seasonal blocks as desired, for example data shifted by ``1
day'', ``2 days'', or ``1 week'', etc. The following code chunk
generates Figure \ref{fig:wrap-ped}.

@@

In response to the slider input, the plot will be updated and loaded
with newly transformed data. Keeping the application as performant as
possible is the top priority. Without completely redrawing the plot, the
\texttt{plotly.js} react method is invoked internally. The underlying
tsibble data is being called back and processed in R. Only transformed
data gets fed back to the shiny server, along with reseting the x-axis
ranges and breaks. The rest plot configurations, such as marks, y-axes,
and layouts, are properly cached.

>>>>

\begin{itemize}
\tightlist
\item
  An ensemble of graphics
\item
  Accelerate the exploratory data visualisation process
\end{itemize}

@@

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduces a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. That said, a \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} classes with temporally
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns.
Below displays the monthly Australian retail trade turnover data
(\texttt{aus\_retail}), available in the \CRANpkg{tsibbledata} package.
The \texttt{Month} column holds year-months as \texttt{index}. The
\texttt{State} together with \texttt{Industry} are the identifiers for
these 152 series, highlighted as \texttt{key}. Note that the column
\texttt{Series\ ID} could be an alternative option for setting up
\texttt{key}, but \texttt{State} and \texttt{Industry} are more readable
and informative. The \texttt{index} and \texttt{key} are ``sticky''
columns to a \texttt{tsibble}, forming critical pieces for fluent
temporal data analysis later.

@@

The R community has a long-standing contribution to interactive
visualisation systems in abundance. These systems can be roughly divided
into two classes: ones with web technology and the others without.

R \CRANpkg{shiny} and \CRANpkg{htmlwidgets} lay a fundamental
infrastructure in marrying R with HTML elements and Javascript for
interactivity, although they do not emphasise on interactive graphics.
The \CRANpkg{htmlwidgets} package opens a door to embedding Javascript
libraries into R such that users are able to write only R code to
generate web-based plots. Many Javascript charting libraries have been
ported to R as HTML widgets, including \CRANpkg{plotly},
\CRANpkg{rbokeh}, and \CRANpkg{leaflet} for maps. To enable interactions
between different widgets, it can be achieved with \CRANpkg{shiny} or
\CRANpkg{crosstalk}. The \CRANpkg{crosstalk} extends
\CRANpkg{htmlwidgets} with shared R6 instances to support linked
brushing and filtering across widgets.

If not using web technology, two examples are provided here:
\CRANpkg{loon} based on Tcl/Tk and \CRANpkg{cranvas} based on Qt. They
offer a wide array of pre-defined interactions, such as selecting and
zooming, to manipulate plots via mouse, keyboard strokes, and menus. The
\CRANpkg{cranvastime} is an add-on of \CRANpkg{cranvas}, which develops
a set of novel interactions for temporal data, for example wrapping and
mirroring.

@@

The \CRANpkg{tsibbletalk} package, inspired by the \CRANpkg{crosstalk}
package, introduces a shared tsibble instance on top of a
\texttt{tsibble} to allow for frictionless communication between
different plots for temporal data. The \texttt{as\_shared\_tsibble()}
function provides an entry point in the integrated flow, turning a
\texttt{tsibble} to a shared instance (i.e.~\texttt{SharedTsibbleData}
subclassing of \texttt{SharedData} from \CRANpkg{crosstalk}) that powers
data transmission across multiple views. The \CRANpkg{tsibbletalk}
package aims to streamline interactive graphical analysis with the focus
of temporal and structured linking.

@@

In summary, shared tsibble data nicely bridges between the
\CRANpkg{crosstalk} and \textbf{tidyverts} ecosystems for temporal data
over their common term ``key''. The \texttt{as\_shared\_tsibble()}
provides a symbolic user interface for effortless construction of a
hybrid of hierarchical and categorical linkings. And the
\texttt{plotly\_key\_tree()} in turn decodes the specification to plot a
tree for data overview and navigation, accompanied with more detailed
plots.

@@

This shiny module, decoupled to \texttt{tsibbleDiceUI()} and
\texttt{tsibbleDiceServer()}, presents a clean interface and forms a
resusable piece in a shiny application. A shiny module provides the
vehicle in modularising shiny applications for both users and
developers. Like all shiny modules, the first argument in both functions
requires a user-supplied session id that must be unique. The UI function
\texttt{tsibbleDiceUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is certainly the server
function \texttt{tsibbleDiceServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, and thereby the time
\texttt{index} can be retrieved for dissection. The \texttt{period}
option semantically takes a desired number of seasonal periods to be
shifted, for example data shifted by ``1 day'', ``2 days'', or ``1
week'', etc. In other words, the \texttt{period} defines the grind
level. As for date-times (represented by \texttt{POSIXt}), it ranges
from from fine ``day'' to coarse ``year''. The following code snippet
generates Figure \ref{fig:wrap-ped}.

@@

In response to the slider input, the plot will be updated and loaded
with newly transformed data. At its core, keeping the application as
performant as possible is the top priority. Without completely redrawing
the plot, the \texttt{plotly.js} react method is invoked internally. The
underlying tsibble data is being called back and processed in R. Only
transformed data gets fed back to the shiny server, along with reseting
the x-axis ranges and breaks. The rest plot configurations, such as
marks, y-axes, and layouts, are properly cached.
